---
title: "Repro_Research_CP2"
author: "Brian Liswell"
date: "12/3/2020"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Repro Course Project 2: Title that Briefly summarizes data analysis

Storms are bad.  If you can't stop them, try to live better with them.

## Synopsis

!!!!!!!!!!!!
Describes and summarizes the data analysis in less than 10 sentences

This project involves exploring the U.S. National Oceanic and Atmospheric Administration's (NOAA) storm database.

Questions
Your data analysis must address the following questions:

**1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?**

The Dataset has columns for "FATALITIES" and "INJURIES".  Those will be the categories for population health.

**2. Across the United States, which types of events have the greatest economic consequences?**

The Dataset has columns for "FATALITIES" and "INJURIES".  Those will be the categories for population health.



### 1. Has either a (1) valid RPubs URL pointing to a data analysis document for this assignment been submitted; or (2) a complete PDF file presenting the data analysis been uploaded?

If you are reading this, the answer is yes.

### 2. Is the document written in English?

I don't want to brag, but I speak English like it's the first language I learned.

### 3. Does the analysis include description and justification for any data transformations?

This question should be at the end.
!!!!!!!!!!!!!!!!!!!!

### 4. Does the document have a title that briefly summarizes the data analysis?
What's the question the analysis/synthesis is trying to answer?
!!!!!!!!!!!!!!!!!!!!

### 5. Does the document have a synopsis that describes and summarizes the data analysis in less than 10 sentences?

See Synopsis at the top.

### 6. Is there a section titled "Data Processing" that describes how the data were loaded into R and processed for analysis?

See below, until question 7.

## Data Processing

### Get the data


```{r echo=TRUE}

file_loc <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
filename <-'Storm_data.csv.bz2'
download.file(file_loc,filename)
data <- read.csv(filename)

ev_names <- data$EVTYPE
dim(data)
dim(ev_names)
```

Most of the columns is needless for this activity. Pair down to just the possibly needed variables.


```{r echo=TRUE}
var_names <- names(data)
names_to_pair <- var_names[c(7,8,22,23,24,25,27)]
# [1] "STATE"      "EVTYPE"     "MAG"        "FATALITIES"
# [5] "INJURIES"   "PROPDMG"    "CROPDMG" 
data <- data[names_to_pair]
```

902297 seperate events recorded, asssigned an Event Type.
985 different "Event" Type names.
Broadly speaking the Events can be categorized into 7 categories.


```{r echo=TRUE}
var_names <- names(data)
names_to_pair <- var_names[c(7,8,22,23,24,25,27)]


storm_names <-c("wind", "tstm", "tornado", "tropical", "storm", "torrent", "thunder", "hurric", "blizzard")
big_wind_names <-c("tornado", "hurric", "gustnado", "funnel")
precip_names <-c("hail", "rain", "snow", "shower", "precip", "sleet")
mud_names <- c("mud")
flood_names <-c("flood","surf","surge","current")
temper_names <- c("cold","heat","hot","blizzard","freez","frost")

```

Add a new Category to data data.frame, based on 7 category names, and "Other" if it doesn't fit.  Use a function, get_tf, to compare different possible names in category.

```{r echo=TRUE}
get_tf <-function(sub_names,ev_names){
        tf <- logical(length(ev_names))
        for(n in sub_names){
                #print(n)
                temp <- grepl(n,tolower(ev_names))
                tf <-tf + temp
        }
        tf <- tf >0  
}
storm_tf <- get_tf(storm_names,ev_names)
big_wind_tf <- get_tf(big_wind_names,ev_names)
precip_tf <- get_tf(precip_names,ev_names)
mud_tf <- get_tf(mud_names,ev_names)
flood_tf <- get_tf(flood_names,ev_names)
temper_tf <- get_tf(temper_names,ev_names)

categ <- character(length(ev_names))
categ[storm_tf] <-"storm"
categ[big_wind_tf] <- "big_wind"
categ[precip_tf] <- "precip"
categ[mud_tf] <- "mud"
categ[flood_tf] <- "flood"
categ[temper_tf] <- "temper"
categ[!nzchar(categ)] <- "other"

data$Broad_categ <- categ
```

Split up the data based on the broad categories.
A "pairs" plot on the original "data" takes forever, and might be useless.
"pairs" plots on the seperate broad categories also take a long time.  Save them to file if you wish.  Entire for loop took more than 45 minutes to make files.

```{r echo=TRUE}
data_sub_categ_split <- split(data,data$Broad_categ)
categ_names <- names(data_sub_categ_split)
# 
# for(n in categ_names){
#         com <-paste0("pairs(data_sub_categ_split$",n,")")
#         plot_name <- paste0("plot_",n,".png")
#         png(file=plot_name)
#         eval(parse(text = com))
#         dev.off()
# }

# pairs(data_sub_categ_split$other)

```

Now sum INJURIES, FATALITIES, PROPDMG and CROPDMG for the 7 subcategories

```{r echo=TRUE}
FATALITIES <- data.frame(Category = categ_names)
FATALITIES$Total <-0
row.names(FATALITIES) <- categ_names
INJURIES <- FATALITIES
PROPDMG <- FATALITIES
CROPDMG <- FATALITIES

for(n in categ_names){
        FATALITIES[n,"Total"] <- sum(data_sub_categ_split[[n]]$FATALITIES)
        INJURIES[n,"Total"] <- sum(data_sub_categ_split[[n]]$INJURIES)
        PROPDMG[n,"Total"] <- sum(data_sub_categ_split[[n]]$PROPDMG)
        CROPDMG[n,"Total"] <- sum(data_sub_categ_split[[n]]$CROPDMG)
}

```


### 7. Is there a section titled "Results" where the main results are presented?

## Results

Create Bar charts for the 4 variables of interest.  Point out interesting things.

```{r echo=TRUE}
library(ggplot2)
p<-ggplot(data=FATALITIES, aes(x=Category, y=Total)) +
  geom_bar(stat="identity")+ggtitle("Fatalities")
print(p)

p<-ggplot(data=INJURIES, aes(x=Category, y=Total)) +
  geom_bar(stat="identity")+ggtitle("INJURIES")
print(p)
p<-ggplot(data=PROPDMG, aes(x=Category, y=Total)) +
  geom_bar(stat="identity")+ggtitle("PROPDMG")
print(p)
p<-ggplot(data=CROPDMG, aes(x=Category, y=Total)) +
  geom_bar(stat="identity")+ggtitle("CROPDMG")
print(p)


```





